Discussion

### 2020-07-27 text/discussion.Rmd
fix bug in references, edit discussion

@@

One of the biggest hurdles in cancer research is the size of the search space;
~20,000 protein-coding genes is comparable with the number of tumor samples,
even with multiple mutations per sample. And that is not accounting for the fact
that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
@tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @non-coding-mutations-Zhang]. Hence, we sought to
simplify the problem by investigating cancer as a disease of basic cellular and
biochemical pathways, which we accomplished by translating gene-level mutations
into pathway level disruptions. Our approach differs from previously proposed
methods [@PATHNET] in that we chose to focus on the pathway as the unit of
disruption instead of the gene, employing the conceit that individual mutations
may be sufficient to disrupt entire pathways (as is often the case with
drivers). Rather than evaluating the burden of mutations on each pathway within
or across tumors, which can be biased by the number of genes or the coverage of
pathway-linked genes in the genome, and which we find to be relatively uniformly
distributed, we scored each pathway in binary fashion, allowing one mutation or
multiple mutations to count equally for each sample. To our knowledge, this
approach has not previously been attempted in the literature in spite of its
relative simplicity. We limited our analysis to mutations with likely
deleterious effects, including missense, nonsense, frame-shift, stop-gain and
splice site mutations in genes that are actively expressed in each cancer type,
thus avoiding bias from transcription coupled repair. Our method of filtration
differs from the "rank-and-cut" method of [@PCAWG-MAIN] but represents a
reasonable attempt to account for the same biases. We also filtered genes on the
basis of pathway membership in biochemical pathways, excluding curated genesets
related to diseases, syndromes, or classes of proteins ( _e.g._ transcription
factors). We chose this approach to limit redundancy, which is already high in
the remaining pathways for some driver genes, and to exclude biologically
unrelated collections of genes. We do not account for non-coding mutations,
amplification events or structural variants, though we hope to include them in
subsequent studies (see below). Our pathway-centric analysis also allows us to
circumvent the problem that some cancer-promoting mutations likely occur at low
enough frequencies as to be indistinguishable from background. Attempts to
capture these associations better using Bayesian networks have yielded important
new insights into this problem and new genes that would not have been possible
with traditional analysis [@hornNetSigNetworkbasedDiscovery2018], supporting the
case that pathway-centric approaches like ours have the potential to increase
statistical power.

The hypothesis that cancer is the result of dysfunction from
a limited number of basic cellular processes common to all living human cells
was first introduced and later expanded on in a pair of classic essays by
Hanahan and Weinberg [@weinberg200; @hallmarks_II]. In this view, all tumors
regardless of tissue of origin acquire 6 essential phenotypes; sustaining
proliferative signaling, evading growth suppressors, replicative immortality,
evading apoptosis, supportive angiogenesis, tissue invasion and metastasis
[@weinberg2000]. Advancements in the field over the ensuing decade led to the
inclusion of new tumor promoting processes, namely deregulation of cellular
energy cycle, tumor promoting inflammation, evasion of the immune system, genome
instability, and on top of these acquired traits a new appreciation for the
concept of the tumor microenvironment and its contribution to the evolution and
promotion of metastatic tumorigenesis [@hallmarks_II].

An alternative hypothesis is that every tumor belongs to one of a large number
of syndromes which are unique to each tissue-of-origin, that share some
mechanisms and treatment strategies. Recent publication of TCGA consortium
papers present a view largely, and surprisingly, consistent with this hypothesis
[@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018].
Perhaps owing to the intractable complexity of genomic, proteomic data and
patient metadata in all its forms, the inescapable conclusion thus far is that
tissue-of-origin remains the most important driver of tumor characteristics at
every scale and by every measure. Our observations contrast with this view, and
instead support an interpretation of publicly available data in which all tumors
manifest one of a limited number of metastable phenotypes resulting from
disruptions in basic pathways represented by our classes (**figure 1**). Each of
these classes may represent unique combinations of pathway disruptions with
increased fitness regardless of tissue of origin, which then converge on end
stage cancer by accumulating complementary mutations to the other classes
(**figures 1L and M** and **Table 2**).

We find that most tumors segregate into one of twelve biologically distinct
clusters, regardless of the cancer type (**figure 1G,I** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
arm clusters 8 and 9, and Kidney Chromophobe in cluster 1, see **supplementary
figure S1**) are unevenly distributed, though we could not identify cancer type
that was exclusive to a single tumor class. Only cancer types with the fewest
samples were found to be absent from one or more classes at all. Somewhat
surprisingly to us, this finding extends to histological subtypes of breast,
head and neck cancers, leukemias, _etc._ (see **Figure 1J** and **Supplemental
Figure S2**). This result implies that many, if not most, histological subtypes
could reflect differences in cell-of-origin, rather than fundamental differences
in proximal cause. The five major subtypes of breast cancer have histological
and molecular correlates that define them and how they respond to experimental
stimuli [@breast-subtypes2]. It has been hypothesized that differences in the
molecular regulators of development in the precursor cell types present in the
breast epithelium drive the various histological phenotypes [@breast-subtypes;
@breast-subtypes3]. Consistent with this view we found that breast tumor samples
of the Luminal A subtype were heavily biased toward membership in clusters 1 and
7, and basal tumor samples were biased toward cluster 6 (**figure 1J**).
However, both subtypes also contained samples in every other class (without
exception), and Luminal B and Her2 positive samples are distributed across
clusters. We interpret these data to mean that inherited cell-of-origin
molecular profiles may predispose certain precursor cell types within the breast
epithelium to forming tumors of one class or another but is not determinative;
this view is compatible with the previously stated hypothesis but opens the way
for a more granular view of individual tumors. It would be surprising if we did
not observe bias for some cancers and subtypes amongst our classes, since some
treatment regimens have greater efficacy for a significant fraction of patients
of a given cancer/histological subtype [@breast-subtypes2]. Nonetheless, the
basis for some tumors being treatment-refractory in spite of receiving the
standard of clinical care for diagnostic markers remains elusive. Doubtless some
of this is due to bad luck, as tumors can metastasize and remain dormant years
before they are detected at distal sites, or resistant clones may have already
arisen at undetectable levels [@hallmarks_II], but our analysis suggests the
possibility of identifying more informative molecular, histological or cellular
subtypes that could form a basis for future separation of patients into
different precision treatment regimens.

It could be that the classes we identified are consistent with a signature of
overall disease progression, such that successive, adjacent classes on the UMAP
projection exhibit increasing specificity. We identified no such trend (see
**figure 1H**) and each class instead was associated with unique combinations of
pathways (as in **figures 2B, C**). In our view, these observations suggest that
cancers have available to them a limited number of stable configurations with
respect to pathway disruption which are consistent with increased fitness and
therefore positive selection.

@@

Finally, we must remark on the limitations of our work exemplified in two of the
classes, 1 and 10, which for which we did not find distinctive associations with
pathways (**figure 2B, C**). Class 1 has a relatively low proportion of mutated
pathways across the board, although it is broadly enriched in many of the same
tumor-promoting pathways common to the other groups. This likely does not
reflect a true difference in stages of tumor evolution, as our data show clearly
that this class is as likely to contain stage IV metastatic tumors as it is to
contain those of stage I (**figure 1L, M**). This class likely represents a
group of tumors with large aberrations in methylation, copy number alterations,
structural variants, or combinations thereof which we excluded from our
analysis. Consistent with this, kidney chromophobe and thyroid cancers have high
proportions of SVs _vs._ other variant types [@PCAWG-MAIN] and are heavily
skewed to cluster 1 membership (**Supplementary figure S1**). Likewise, Class 8
represents a group of hyper-mutated tumors that harbor so many mutations that
virtually no pathway is unaffected. It seems likely that a significant fraction
of the samples flagged as “mutant” for each pathway are actually unaffected,
consistent with those samples have low information content. This could be
addressed with application of ever-increasingly sophisticated bioinformatics
algorithms for filtering likely passenger mutations [ _e.g._ @spliceAI;
@deep-learning-clinical-variants]. In the future, we hope to incorporate these
other data into a pathway-centric analysis as we have done here for point
mutations and indels.

@@

By modeling CDC longevity data as a baseline risk function we showed
that each of our pathway disruption classes exhibit cancer-type specific effects
on expected survival (**figure 4**). Contrary to what one might expect at the
outset, we found that models assuming different survival based on class _and_
cancer type fit better than models in which these variables are independent
factors. However, considering that within each cancer type there are different
clinical standards of care, and even within classes of drugs the preferred
treatment can very between cancers, it makes sense in hindsight that they should
all be different. Contrast the situation with ovarian vs breast cancer, which
are both hormonally driven cancers, for example. Ovarian cancer has but one main
treatment, platinum, whereas breast cancer patients have a variety of treatment
regimens based on molecular subtype and other factors. It is easy to see how
standard of care for one cancer and class could produce different outcomes than
the same class of a different cancer type under the circumstances.
Unfortunately, given the diversity drug classes and treatments (including
non-chemical treatments such as surgery and palliative care), we lack sufficient
power in the current study to explore these variables in the TCGA data. It is
our hope that more focused future studies will help to distinguish between
treatment-specific effects or survival given different pathway disruption
classes.

@@

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, class-specific
mutations and metastasis. General cancer promoting processes must include
factors that relate to genome stability and immortality, as "enabling
characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
disrupted in all or most of the defined classes (**supplementary figure S1**)
and often are the result of aberrations involving common driver genes such as
BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._
(**supplementary figure S3** and **supplemental file 1**). Class-specific
evolution must involve the acquisition of combinatorial disruptions to pathways
that may individually be harmful but together produce more specialized cancer
phenotypes and increased fitness (as in **figure 4** and **tables 1** and
**ST2**). Importantly, our work does not distinguish between the order in which
these mutations should accumulate. This could be addressed in a future study by
integrating our analysis with evolutionary analysis of clonality, drawing
inferences from variant allele frequencies as in @PCAWG-EVO. However, since many
of the genes in the non-cluster-specific pathways involve genome stability, it
is reasonable to hypothesize that such mutations promote the acquisition of
class-specific defects via random mutation and natural selection, thus producing
the classes we observed. Indeed, a subset of known oncogenic driver mutations
are highly enriched in early arising clones, whereas later arising clones have
much greater diversity in driver mutations [@PCAWG-EVO]. Moreover, driver genes
that are known to be responsible for discrete mutation signatures such as
APOBEC, BRCA1 and BRCA2 produce mutational hotspots reflecting the varying
selective pressures in different tissues [@PCAWG-MAIN].

One of the drawbacks of bulk tumor whole genome sequencing data
is the problem of tumor heterogeneity, thus consortium samples are likely to
contain contamination from support tissue and inflammatory and immune cells of
the innate and adaptive immune systems, potentially harboring cancer supporting
mutations [@bhowmick_review]. We think it will be instructive to explore these
ideas in the context of single cell experiments.

@@

We report that pathway enrichment in metastatic tumors
across all classes (pooled) yielded generally lower effect sizes and larger
p-values than the cluster specific analysis, suggesting that signal is diluted
when the samples are analyzed this way, and supporting the view that metastasis
has class-specific requirements. Since the number of metastatic tumors is
relatively low this part of our analysis is likely underpowered and subject to
expanded analysis with larger cohorts. Nonetheless, the class-specific
metastatic pathway enrichment we observe (in **table 2**) could be a result of
the pathways directly promoting metastatic phenotypes in these classes, or
because they acquire a stronger or more complete oncogenic phenotype, since we
don’t know the order in which they were acquired in individual tumor samples.
The fact that most metastatic enrichment is cluster-specific and has a tendency
to overlap with cluster-specific pathways from non-metastatic tumors of
neighboring clusters favors the latter interpretation, and suggests a mechanism
of complementarity, in which newly acquired mutations result in similar classes
converging on a single deadly phenotype that captures most or all of the
critical features of end-stage cancer. Some clusters were characterized by
cluster-specific pathways that seem superficially unrelated; for example cluster
3 has mutations impacting protein translation and turnover, whereas cluster 5
has defects in RNA metabolism and clusters 7, 8 and 9 have defects in the more
traditional extra-cellular signaling pathways (as exemplified in **table 1**).
It may be that the primary effects of mutations in protein and RNA metabolism in
clusters 3 and 5 point to currently unappreciated mechanisms impacting hallmark
pathways. Clusters 3 and 5 are also marked by a decidedly different set of
pathways in metastasis. Clusters 3 and 5 notably have (in addition to signaling
pathways) mutations in glucagon and glycogen metabolism, respectively, in their
metastases (**table 2**), consistent with energy metabolism as one of the new
hallmarks [@hallmarks_II], but also distinguishing these classes from the others
in the order or nature of hallmark acquisition. Thus, even at relatively low
power, our analysis of metastasis further supports pathway-centric analysis of
tumors as indicative of real differences in underlying biology. As a major
caveat to this, we acknowledge that many therapies do target specific signaling
pathways ( _e.g._ EGFR or HER2), but understanding the broader context of the
genetic background and pathway vulnerability of tumors containing such markers
may aid in creating smarter combination therapies. We submit that when we
discover the requirements of each cluster with respect to pathway disruptions
and metastasis we may be able to target them therapeutically and prevent further
adaptation.

>>>>

---
title: "Discussion"
output: 
  bookdown::html_document2:
    fig_captions: TRUE
    number_sections: FALSE
    toc: FALSE
    code_folding: hide
    theme: flatly
    smart: true
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

@@

One of the biggest hurdles in cancer research is the size of the search space;
~20,000 protein-coding genes is comparable with the number of tumor samples,
even with multiple mutations per sample. And that is not accounting for the fact
that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
@tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @non-coding-mutations-Zhang]. Hence, we sought to
simplify the problem by investigating cancer as a disease of basic cellular and
biochemical pathways, which we accomplished by translating gene-level mutations
into pathway level disruptions. Our approach differs from previously proposed
methods [@PATHNET] in that we chose to focus on the pathway as the unit of
disruption instead of the gene, employing the conceit that individual mutations
may be sufficient to disrupt entire pathways (as is often the case with
drivers). Rather than evaluating the burden of mutations on each pathway within
or across tumors, which can be biased by the number of genes or the coverage of
pathway-linked genes in the genome, and which we find to be relatively uniformly
distributed, we scored each pathway in binary fashion, allowing one mutation or
multiple mutations to count equally for each sample. To our knowledge, this
approach has not previously been attempted in the literature in spite of its
relative simplicity. We limited our analysis to mutations with likely
deleterious effects, including missense, nonsense, frame-shift, stop-gain and
splice site mutations in genes that are actively expressed in each cancer type,
thus avoiding bias from transcription coupled repair. Our method of filtration
differs from the "rank-and-cut" method of [@PCAWG-MAIN] but represents a
reasonable attempt to account for the same biases. We also filtered genes on the
basis of pathway membership in biochemical pathways, excluding curated genesets
related to diseases, syndromes, or classes of proteins ( _e.g._ transcription
factors). We chose this approach to limit redundancy, which is already high in
the remaining pathways for some driver genes, and to exclude biologically
unrelated collections of genes. We do not account for non-coding mutations,
amplification events or structural variants, though we hope to include them in
subsequent studies (see below). Our pathway-centric analysis also allows us to
circumvent the problem that some cancer-promoting mutations likely occur at low
enough frequencies as to be indistinguishable from background. Previous attempts
to capture these associations better using Bayesian networks have yielded
important new insights into this problem and new genes
[@hornNetSigNetworkbasedDiscovery2018] that would not have been possible with
traditional analysis, supporting the case that pathway-centric approaches like
ours have the potential to increase statistical power.

The hypothesis that cancer is the result of dysfunction from
a limited number of basic cellular processes common to all living human cells
was first introduced and later expanded on in a pair of classic essays by
Hanahan and Weinberg [@weinberg200; @hallmarks_II]. In this view, all tumors
regardless of tissue of origin acquire six essential phenotypes; sustaining
proliferative signaling, evading growth suppressors, replicative immortality,
evading apoptosis, supportive angiogenesis, tissue invasion and metastasis
[@weinberg2000]. Advancements in the field over the ensuing decade led to the
inclusion of new tumor promoting processes, namely deregulation of cellular
energy cycle, tumor promoting inflammation, evasion of the immune system, genome
instability, and on top of these acquired traits a new appreciation for the
concept of the tumor microenvironment and its contribution to the evolution and
promotion of metastatic tumorigenesis [@hallmarks_II].

An alternative hypothesis is that every tumor belongs to one of a large number
of syndromes which are unique to each tissue-of-origin, that share some
mechanisms and treatment strategies. Recent publication of TCGA consortium
papers present a view largely, and surprisingly, consistent with this hypothesis
[@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018].
Perhaps owing to the intractable complexity of genomics, proteomics, and
patient metadata in all its forms, the inescapable conclusion thus far is that
tissue-of-origin remains the most important driver of tumor characteristics at
every scale and by every measure. Our observations contrast with this view, and
instead support an interpretation of publicly available data in which all tumors
manifest one of a limited number of metastable phenotypes resulting from
disruptions in basic pathways represented by our classes (**figure 1**).

We attempted to account for our clusters in terms of more trivial explanations.
For example, it could be that the classes we identified are consistent with a
signature of overall disease progression, such that successive, adjacent classes
on the UMAP projection exhibit increasing specificity. We identified no such
trend in the number of mutations (see **figure 1H**), the relative staging
(**figure 1L**) or metastasis (**figure 1M**); and each class instead was
associated with unique combinations of pathways (as in **figures 2B, C**).

We find that most tumors segregate into one of twelve biologically distinct
clusters, regardless of the cancer type (**figure 1G,I** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
mutant clusters 8 and 9, and Kidney Chromophobe in cluster 1, see **supplementary
figure S1**) are unevenly distributed, though we could not identify any cancers
that were exclusive to a single tumor class. Only cancer types with the fewest
samples were found to be absent from one or more classes at all. Somewhat
surprisingly to us, this finding extends to histological subtypes of breast,
head and neck cancers, leukemias, _etc._ (see **Figure 1J** and **Supplemental
Figure S2**). This result implies that many, if not most, histological subtypes
could reflect differences in cell-of-origin, rather than fundamental differences
in proximal cause. The four major subtypes of breast cancer correspond to histological
and molecular expression profiles that define them and how they respond to experimental
stimuli [@breast-subtypes2]. It has been hypothesized that differences in the
molecular regulators of development in the precursor cell types present in the
breast epithelium drive the various histological phenotypes [@breast-subtypes;
@breast-subtypes3]. Consistent with this view we found that breast tumor samples
of the Luminal A subtype were heavily biased toward membership in clusters 1 and
7, and basal tumor samples were biased toward cluster 6 (**figure 1J**).
However, both subtypes also contained samples in every other class (without
exception), and Luminal B and Her2 positive samples are distributed across
clusters. We interpret these data to mean that inherited cell-of-origin
molecular profiles may predispose certain precursor cell types within the breast
epithelium to forming tumors of one class or another but is not determinative;
this view is compatible with the previously stated hypothesis but opens the way
for a more granular view of individual tumors. It would be surprising if we did
not observe bias for some cancers and subtypes amongst our classes, since some
treatment regimens have greater efficacy for patients
of a given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the
basis for some tumors being treatment-refractory in spite of receiving the
standard of clinical care for diagnostic markers remains elusive. Doubtless some
of this is due to bad luck, as tumors can metastasize and remain dormant years
before they are detected at distal sites, or resistant clones may have already
arisen at undetectable levels [@hallmarks_II], but our analysis suggests the
possibility of identifying more informative molecular, histological or cellular
subtypes that could form a basis for future separation of patients into
different precision treatment regimens.

@@

Finally, we must remark on the limitations of our work exemplified in two of the
classes, 1 and 10, which for which we did not find distinctive associations with
pathways (**figure 2B, C**). Class 1 has a relatively low proportion of mutated
pathways across the board, although it is broadly enriched in many of the same
tumor-promoting pathways common to the other groups. This likely does not
reflect a true difference in stages of tumor evolution, as our data show clearly
that this class is as likely to contain stage IV metastatic tumors as it is to
contain those of stage I (**figure 1L, M**). This class likely represents a
group of tumors with aberrations in methylation, copy number alterations,
structural variants, or combinations thereof which we excluded from our
analysis. Consistent with this, kidney chromophobe and thyroid cancers have high
proportions of SVs _vs._ other variant types [@PCAWG-MAIN] and are heavily
skewed to cluster 1 membership (**Supplementary figure S1**). Likewise, Class 8
represents a group of hyper-mutated tumors that harbor so many mutations that
virtually no pathway is unaffected. It seems likely that a significant fraction
of the samples flagged as “mutant” for each pathway are actually unaffected,
consistent with those samples have low information content. This could be
addressed with application of ever-increasingly sophisticated bioinformatics
algorithms for filtering likely passenger mutations [ _e.g._ @spliceAI;
@deep-learning-clinical-variants]. In the future, we hope to incorporate these
other data into a comprehensive pathway-centric analysis as we have done here
for point mutations and indels.

@@

By modeling CDC longevity data as a baseline risk function we showed that each
of our pathway disruption classes exhibit cancer-type specific effects on
expected survival (**figure 4**). Contrary to what one might expect at the
outset, we found that models assuming different survival based on class _and_
cancer type fit better than models in which these variables are independent
factors. However, considering that within each cancer type there are different
clinical standards of care, and even within classes of drugs the preferred
treatment can very between cancers, it makes sense in hindsight that they should
all be different. Contrast the situation with ovarian _vs._ breast cancer, which
are both hormonally driven cancers, for example. Ovarian cancer has but one main
treatment axis, platinum, whereas breast cancer patients have a variety of
treatment regimens based on molecular subtype and other factors. It is easy to
see how standard of care for one cancer and class could produce different
outcomes than the same class of a different cancer type. Unfortunately, given
the diversity drug classes and treatments (including non-chemical treatments
such as surgery and palliative care), we lack sufficient power in the current
study to explore these variables in the TCGA data. It is our hope that more
focused future studies will help to distinguish between treatment-specific
effects or survival given different pathway disruption classes.

@@

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, class-specific
mutations and metastasis. General cancer promoting processes must include
factors that relate to genome stability and immortality, as "enabling
characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
disrupted in all or most of the defined classes (**supplementary figure S1**)
and often are the result of aberrations involving common driver genes such as
BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._
(**supplementary figure S3** and **supplemental file 1**). Class-specific
evolution must involve the acquisition of disruptions to pathways that may
individually be harmful ( _e.g._ highly proliferative cells are more likely to
senesce) but together produce more specialized cancer phenotype and increased
fitness (as in **figure 4** and **tables 1** and **ST2**). Importantly, our work
does not distinguish between the order in which these mutations should
accumulate. This could be addressed in a future study by integrating our
analysis with evolutionary analysis of clonality, drawing inferences from
variant allele frequencies as in @PCAWG-EVO. However, since many of the genes in
the non-cluster-specific pathways involve genome stability, it is reasonable to
hypothesize that these mutations promote or enable the acquisition of
class-specific defects via random mutation and natural selection, thus producing
the classes we observed. In support of this, the pan-cancer analysis of whole
genomes consortium (PCAWG) found that a subset of known oncogenic driver
mutations are highly enriched in early arising clones, whereas later arising
clones have much greater diversity in driver mutations [@PCAWG-EVO]. Moreover,
driver genes that are known to be responsible for discrete mutation signatures
such as APOBEC, BRCA1 and BRCA2 produce mutational hotspots reflecting the
varying selective pressures in different tissues [@PCAWG-MAIN].

One of the drawbacks of bulk tumor whole genome sequencing data is the problem
of tumor heterogeneity. Consortium samples are likely to contain contamination
from support tissue and inflammatory and immune cells of the innate and adaptive
immune systems, potentially harboring cancer supporting mutations
[@bhowmick_review]. We think it will be instructive to explore these ideas in
the context of single cell experiments.

@@

We report that pathway enrichment in metastatic tumors across all classes
(pooled) yielded generally lower effect sizes and larger p-values than the
cluster specific analysis, suggesting that signal is diluted when the samples
are analyzed this way, and supporting the view that metastasis has
class-specific requirements. Since the number of metastatic tumors is relatively
low this part of our analysis is likely underpowered and subject to expanded
analysis with larger cohorts. Nonetheless, the class-specific metastatic pathway
enrichment we observe (in **table 2**) could be a result of the pathways
directly promoting metastatic phenotypes in these classes, or because they
acquire a stronger or more complete oncogenic phenotype, as we don’t know the
order in which they were acquired in individual tumor samples. The fact that
most metastatic enrichment is cluster-specific and has a tendency to overlap
with cluster-specific pathways from non-metastatic tumors of neighboring
clusters favors the latter interpretation, and suggests a mechanism of
complementarity, in which newly acquired mutations result in similar classes
converging on a single deadly phenotype that captures most or all of the
critical features of end-stage cancer. One explanation for this could be that
tumors must accumulate mutations in a certain order or face elimination due to
any of the selective bottlenecks they face (senescence, the immune system,
_etc._) [@hallmarks_II]. Some clusters were characterized by cluster-specific
pathways that seem superficially unrelated; for example cluster 3 has mutations
impacting protein translation and turnover, whereas cluster 5 has defects in RNA
metabolism and clusters 7, 8 and 9 have defects in the more traditional
extra-cellular signaling pathways (as exemplified in **table 1**). It may be
that the primary effects of mutations in protein and RNA metabolism in clusters
3 and 5 point to currently unappreciated mechanisms impacting hallmark pathways.
Clusters 3 and 5 are also marked by a decidedly different set of pathways in
metastasis. Clusters 3 and 5 notably have (in addition to signaling pathways)
mutations in glucagon and glycogen metabolism, respectively, in their metastases
(**table 2**), consistent with energy metabolism as one of the new hallmarks
[@hallmarks_II], but also distinguishing these classes from the others in the
order or nature of hallmark acquisition. Thus, even at relatively low power, our
analysis of metastasis further supports pathway-centric analysis of tumors as
indicative of real differences in underlying biology. As a major caveat to this,
we acknowledge that many therapies _do_ target specific signaling pathways (
_e.g._ EGFR or HER2), but understanding the broader context of the genetic
background and pathway vulnerability of tumors containing such markers may aid
in creating smarter combination therapies. We submit that when we discover the
requirements of each cluster with respect to pathway disruptions and metastasis
we may be able to target them therapeutically and prevent further adaptation.
